---
title: "Benchmarking proteomics"
author: "Ema Svetlicic"
date: "2024-04-21"
output: html_document
---




# 1. Import datasets.

The datasets are output from 4 different softwares used (MaxQuant, Proteome Discoverer, DIA-NN and Spectronaut). They were pre-processed with custom script to make them all have the same format (dfWide_xx.txt)

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


```{r}
library(ggplot2)
library(dplyr)
library(tidyr)
library(stringr)
library(limma)


dfWide_MQ <- read.delim(
  "dfWide_MQ.txt",
  sep = " ",
  header = TRUE,
  check.names = FALSE,
  stringsAsFactors = FALSE
)

dfWide_SN <- read.delim(
  "dfWide_SN.txt",
  sep = " ",
  header = TRUE,
  check.names = FALSE,
  stringsAsFactors = FALSE
)

dfWide_diann <- read.delim(
  "dfWide_diann.txt",
  sep = " ",
  header = TRUE,
  check.names = FALSE,
  stringsAsFactors = FALSE
)

dfWide_PD <- read.delim(
  "dfWide_PD.txt",
  sep = " ",
  header = TRUE,
  check.names = FALSE,
  stringsAsFactors = FALSE
)

dfWide_TMT<-read.delim("dfWide_TMT.txt",
  sep=" ",
   header=TRUE,
   check.names=FALSE,
   stringsAsFactors=FALSE
)


ED<-read.delim("Bench/experimental_design.txt",
  sep="\t",
   header=TRUE,
   check.names=FALSE,
   stringsAsFactors=FALSE
)

```



# 2. Create a list of dfWides

```{r}

df_list <- list(
  DIA_NN = dfWide_diann,
  MQ = dfWide_MQ,
  PD = dfWide_PD,
  SN = dfWide_SN,
  TMT= dfWide_TMT
)

```

# 3. Load the annotation file (which protein belong to S. aureus or H. sapiens)

```{r}
annot<-read.delim("annotation_file.txt", header=TRUE)


library(dplyr)
library(tidyr)
library(ggplot2)


annot_category <- annot %>%
  mutate(
    Category = case_when(
      grepl("Homo sapiens", Organism) ~ "HS",
      grepl("Staphylococcus aureus", Organism) ~ "SA",
      TRUE ~ NA_character_
    )
  ) %>%
  select(Entry, Category)



#  FUNCTION TO CALCULATE HS AND SA COUNTS

get_counts <- function(df, method) {
  
  df_long <- df %>%
    tibble::rownames_to_column("Entry") %>%
    pivot_longer(
      cols = -Entry,
      names_to = "Sample",
      values_to = "Value"
    ) %>%
    left_join(
      annot_category %>%
        distinct(Entry, .keep_all = TRUE),
      by = "Entry"
    )
  
  counts <- df_long %>%
    filter(Category %in% c("HS", "SA")) %>%
    group_by(Sample, Category) %>%
    summarise(
      Count = n_distinct(Entry[!is.na(Value)]),
      .groups = "drop"
    ) %>%
    mutate(Method = method)
  
  return(counts)
}


count_results <- lapply(
  names(df_list),
  function(x) {
    get_counts(
      df = df_list[[x]],
      method = x
    )
  }
)

all_counts <- bind_rows(count_results)



# PLOT THE NUMBER OF IDENTFIED PROTEINS FOR HS AND SA

ggplot(
  all_counts,
  aes(
    x = Method,
    y = Count,
    fill = Category
  )
) +
  geom_boxplot(
    position = position_dodge(width = 0.75),
    width = 0.6,
    outlier.shape = NA
  ) +
  geom_jitter(
    aes(group = Category),
    position = position_jitterdodge(
      jitter.width = 0.15,
      dodge.width = 0.75
    ),
    size = 1.5,
    alpha = 0.6
  ) +
  scale_fill_manual(
    values = c(
      "HS" = "royalblue3",
      "SA" = "coral2"
    )
  ) +
  labs(
    x = "Quantification method",
    y = "Number of quantified proteins",
    fill = "Organism"
  ) +
  theme_bw() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    legend.position = "top",
    panel.grid.major.x = element_blank()
  )

```

# 4. Explore missing values in the datasets

```{r}
library(naniar)

missing_long <- lapply(names(df_list), function(method) {

  df <- df_list[[method]]

  df %>%
    mutate(Accession = rownames(df)) %>%
    pivot_longer(
      cols = -Accession,
      names_to = "Sample",
      values_to = "Value"
    ) %>%
    mutate(
      Missing = is.na(Value),
      Method = method
    )

}) %>%
  bind_rows()


ggplot(
  missing_long,
  aes(x = Sample, y = Accession, fill = Missing)
) +
  geom_raster() +
  facet_wrap(~Method, scales = "free_y") +
  scale_fill_manual(
    values = c(
      "FALSE" = "coral",
      "TRUE" = "black"
    ),
    labels = c(
      "FALSE" = "Present",
      "TRUE" = "Missing"
    )
  ) +
  labs(
    x = "Sample",
    y = "Protein",
    fill = ""
  ) +
  theme_minimal() +
  theme(
    axis.text.y = element_blank(),
    axis.ticks.y = element_blank(),
    axis.text.x = element_text(
      angle = 90,
      hjust = 1
    ),
    panel.grid = element_blank(),
    strip.text = element_text(face = "bold")
  )
```

TMT does not have missing values since the samples are multiplexed

# 5. Plot the PCA plots

```{r}
library(dplyr)
library(ggplot2)
library(patchwork)


# PCA function


run_pca <- function(df, method, ED) {


  filtered_dfWide <- na.omit(df)

  # prepare annotations
  condition <- ED$Group

  pca_data <- prcomp(t(filtered_dfWide))

  pca_data_perc <- round(
    100 * pca_data$sdev^2 / sum(pca_data$sdev^2),
    1
  )

  df_pca_data <- data.frame(
    PC1 = pca_data$x[,1],
    PC2 = pca_data$x[,2],
    sample = colnames(filtered_dfWide),
    condition = condition
  )



  p <- ggplot(
    df_pca_data,
    aes(PC1, PC2, color = condition)
  ) +
    geom_point(size = 3) +
    theme_bw() +
    stat_ellipse(
      geom = "polygon",
      aes(fill = condition),
      alpha = 0.05,
      show.legend = FALSE,
      level = 0.80
    ) +
    labs(
      title = method,
      x = paste0("PC1  ", pca_data_perc[1], " %"),
      y = paste0("PC2  ", pca_data_perc[2], " %")
    )

  return(p)
}


pca_plots <- lapply(
  names(df_list),
  function(method) {

    run_pca(
      df = df_list[[method]],
      method = method,
      ED = ED
    )

  }
)

names(pca_plots) <- names(df_list)
```

```{r}


pca_combined <- wrap_plots(
  pca_plots,
  ncol = 2
)

pca_combined
```



# 6. Plot coefficient of variation for the proteins 
Assess the reproducibility and dispersion of protein quantification across the different quantification methods.


```{r}


library(dplyr)
library(tidyr)
library(ggplot2)

# ---------------------------------------------------------
# CV calculation for all 5 methods
# ---------------------------------------------------------

method_names <- c(
  "PD"     = "PD",
  "MQ"     = "MQ",
  "TMT"    = "TMT",
  "DIA_NN" = "DIA-NN",
  "SN"     = "SN"
)

cv_results <- lapply(names(df_list), function(method) {
  
  df <- df_list[[method]]
  
  # Convert wide format to long format
  df_long <- df %>%
    tibble::rownames_to_column(var = "Accession") %>%
    pivot_longer(
      cols = -Accession,
      names_to = "Name",
      values_to = "Abundance"
    )
  
  # Add experimental design
  df_long <- df_long %>%
    left_join(
      ED %>% select(Name, Group),
      by = "Name"
    )
  
  # Calculate CV for each protein within each group
  cv <- df_long %>%
    group_by(Accession, Group) %>%
    summarise(
      CV = (
        sd(Abundance, na.rm = TRUE) /
        mean(Abundance, na.rm = TRUE)
      ) * 100,
      .groups = "drop"
    ) %>%
    mutate(
      Method = method_names[method]
    )
  
  cv
})

# ---------------------------------------------------------
# Combine all 5 methods
# ---------------------------------------------------------

cv_data_combined <- bind_rows(cv_results)

# Set method order
cv_data_combined$Method <- factor(
  cv_data_combined$Method,
  levels = c(
    "PD",
    "MQ",
    "TMT",
    "DIA-NN",
    "SN"
  )
)

# Remove invalid CV values
cv_data_combined <- cv_data_combined %>%
  filter(
    is.finite(CV),
    !is.na(Group)
  )

# ---------------------------------------------------------
# PLOT
# ---------------------------------------------------------

p_CV <- ggplot(
  cv_data_combined,
  aes(
    x = Group,
    y = CV,
    fill = Method
  )
) +
  
  geom_violin(
    alpha = 0.5,
    position = position_dodge(width = 0.9),
    trim = TRUE
  ) +
  
  geom_boxplot(
    width = 0.12,
    position = position_dodge(width = 0.9),
    outlier.size = 0.5
  ) +
  
  labs(
    title = "Coefficient of Variation across quantification methods",
    x = "Analysis",
    y = "CV (%)",
    fill = "Method"
  ) +
  
  theme_minimal() +
  
  scale_fill_manual(
    values = c(
      "PD" = "steelblue3",
      "MQ" = "tomato",
      "TMT" = "forestgreen",
      "DIA-NN" = "gold",
      "SN" = "orchid"
    )
  )

p_CV


```



# 7. Run differential expression analysis 
 To estimate quantification accuracy of the tested analysis methods


```{r}
run_DE <- function(df, method) {
  

  df <- as.data.frame(df)
  
  df[] <- lapply(df, as.numeric)
  

  
  dfWide_log <- log2(df)
  
  
  # Remove rows containing only NA
  dfWide_log <- dfWide_log[
    rowSums(!is.na(dfWide_log)) > 0,
    ,
    drop = FALSE
  ]
  

  
  condition <- as.factor(ED$Group)
  
  cond <- as.factor(condition)
  
  design <- model.matrix(~0 + cond)
  
  colnames(design) <- gsub(
    "cond",
    "",
    colnames(design)
  )
  
  

  
  if (ncol(dfWide_log) != nrow(ED)) {
    stop(
      paste(
        method,
        ": number of samples in dataframe does not match ED"
      )
    )
  }
  

# Contrasts

  x <- c(
    "B-A",
    "C-A",
    "D-A",
    "D-B",
    "B-D",
    "D-C"
  )
  
  contrast <- makeContrasts(
    contrasts = x,
    levels = design
  )
  
  

  
  fit1 <- lmFit(
    dfWide_log,
    design
  )
  
  fit2 <- contrasts.fit(
    fit1,
    contrasts = contrast
  )
  
  fit3 <- eBayes(fit2)
  
  

# Extract comparisons

  
  T1 <- topTable(
    fit3,
    coef = 1,
    n = Inf
  )
  
  T2 <- topTable(
    fit3,
    coef = 2,
    n = Inf
  )
  
  T3 <- topTable(
    fit3,
    coef = 3,
    n = Inf
  )
  
  

# Add accession from rownames

  
  T1$Accession <- rownames(T1)
  T2$Accession <- rownames(T2)
  T3$Accession <- rownames(T3)
  
  

# Add annotation

  
  T1 <- T1 %>%
    left_join(
      annot_category,
      by = c("Accession" = "Entry")
    )
  
  T2 <- T2 %>%
    left_join(
      annot_category,
      by = c("Accession" = "Entry")
    )
  
  T3 <- T3 %>%
    left_join(
      annot_category,
      by = c("Accession" = "Entry")
    )
  
  

  T1 <- T1 %>%
    filter(Category %in% c("HS", "SA"))
  
  T2 <- T2 %>%
    filter(Category %in% c("HS", "SA"))
  
  T3 <- T3 %>%
    filter(Category %in% c("HS", "SA"))
  
  

# Combine comparisons

  
  result <- bind_rows(
    
    T1 %>%
      select(
        Accession,
        Category,
        logFC,
        P.Value,
        adj.P.Val
      ) %>%
      mutate(
        Comparison = "B-A",
        Method = method
      ),
    
    T2 %>%
      select(
        Accession,
        Category,
        logFC,
        P.Value,
        adj.P.Val
      ) %>%
      mutate(
        Comparison = "C-A",
        Method = method
      ),
    
    T3 %>%
      select(
        Accession,
        Category,
        logFC,
        P.Value,
        adj.P.Val
      ) %>%
      mutate(
        Comparison = "D-A",
        Method = method
      )
  )
  
  
  return(result)
}
```

```{r}
DE_results <- lapply(
  names(df_list),
  function(x) {
    
    run_DE(
      df = df_list[[x]],
      method = x
    )
    
  }
)

names(DE_results) <- names(df_list)



DE_results <- lapply(
  names(df_list),
  function(x) {
    run_DE(df_list[[x]], x)
  }
)

names(DE_results) <- names(df_list)

DE_all <- bind_rows(DE_results)
```


# 8. Plot logFC and compare to the expected values (theoretical)

```{r}
thresholds <- data.frame(
  Category = c(
    "HS", "HS", "HS",
    "SA", "SA", "SA"
  ),
  Comparison = c(
    "B-A", "C-A", "D-A",
    "B-A", "C-A", "D-A"
  ),
  yintercept = c(
    -0.59, -1.58, -2.45,
     0.42, 0.69, 0.85
  ),
  line_colour = c(
    "blue", "green", "red",
    "blue", "green", "red"
  )
)


p_logFC <- ggplot(
  DE_all,
  aes(
    x = Method,
    y = logFC,
    fill = Method
  )
) +

  geom_boxplot(
    outlier.size = 0.3
  ) +

  facet_grid(
    Category ~ Comparison
  ) +

 

  # Your HS and SA reference lines
  geom_hline(
    data = thresholds,
    aes(
      yintercept = yintercept,
      color = line_colour
    ),
    linetype = "dashed",
    inherit.aes = FALSE
  ) +

  scale_color_identity() +

  labs(
    title = "Protein log2 fold changes across quantification methods",
    x = "Method",
    y = "log2 fold change"
  ) +

  theme_bw() +

  theme(
    axis.text.x = element_text(
      angle = 45,
      hjust = 1
    ),
    legend.position = "none"
  )


p_logFC
```

# 9. Plot the number of proteins with significant and non-significant p-values
To assess the consistency and performance of the different quantification methods.
Ideally, all proteins should show statistically significant p-values.

```{r}


pval_plot_data <- DE_all %>%
  mutate(
    Significance = ifelse(
      adj.P.Val < 0.05,
      "Significant",
      "Not significant"
    )
  ) %>%
  group_by(Method, Comparison, Significance) %>%
  summarise(
    N = n(),
    .groups = "drop"
  ) %>%
  group_by(Method, Comparison) %>%
  mutate(
    Percentage = 100 * N / sum(N)
  ) %>%
  ungroup()

# Set method order
pval_plot_data$Method <- factor(
  pval_plot_data$Method,
  levels = c("PD", "MQ", "TMT", "DIA_NN", "SN")
)

# Plot
p_pvalues <- ggplot(
  pval_plot_data,
  aes(
    x = Method,
    y = Percentage,
    fill = Significance
  )
) +
  geom_col(
    position = "stack",
    width = 0.7
  ) +
  facet_wrap(
    ~ Comparison
  ) +
  labs(
    title = "Statistical significance across quantification methods",
    x = "Quantification method",
    y = "Proteins (%)",
    fill = "P-value"
  ) +
  scale_fill_manual(
    values = c(
      "Significant" = "firebrick3",
      "Not significant" = "grey80"
    )
  ) +
  theme_bw() +
  theme(
    axis.text.x = element_text(
      angle = 45,
      hjust = 1
    ),
    strip.background = element_rect(
      fill = "grey90"
    )
  )

p_pvalues
```

NA : adjusted p-values for proteins where there isn't enough valid quantitative information to estimate the statistics.
